Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
pass up to 2.4

\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part. Let us detail briefly the memory used by the production VM, on the top of the figure. On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.

@@

All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add flexibility through polymorphism. For example, the \texttt{AbstractCompactor} class has two subclasses, one implementing a sweep algorithm and the other a compact one. For production, at Slang-to-C compilation time, all the code is compiled in a single C file. All the flexibility is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm. No polymorphism is available at runtime. However, since polymorphism is available in the simulator, it can be reused for debugging purposes. Still in our example, the \texttt{AbstractCompactor} class has also simulation specific subclasses. Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints. They also keep specific values live so they can be accessed at debugging time.

@@

In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of the native code it generates. The JIT itself is written in Slang and simulated wi th the Smalltalk execution model. To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely. All but the MIPSLE simulators are accessed through Smalltalk primitives.  These are effectively a combination of an FFI call to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. The next two sections describe in detail how this scheme is used to ensure that simulation and production machine code agree, how simulation machine code is interfaced to the rest of the simulation, and the debugging facilities for machine code that are available.

@@

A key requirement to enable effective development of the JIT is that the code generated during simulation be as close as possible to the code generated by the production JIT.  The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation. But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.  The technique we use, involving illegal addresses, also allows us to interface the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and jit, and variables within them, with machine code. To explain this let's consider how the interpreter references a stack frame.

The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code. When machine code wants to enter the run time, for example to have the run time look up a send, or to invoke the scavenger when eden is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine in the run time. The run time can then access the current machine code stack frame though these variables.  In the production VM, the stackPointer and framePointer variables are C variables at known addresses in Data, and routines in the runtime have been compiled from C and exit in Text.  Referencing these variables and invoking runtime routines from machine code is therefore straight-forward.  But in the simulation they are inaccessible.  First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.

%NOTE: I almost only commented part of what Eliot wrote but did not change anything. I basically removed implementation specific details.
The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects. As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction \footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.  %Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  
The processor simulator is invoked in the context of an exception handler for ProcessorException exceptions which invokes %a Smalltalk method, handleSimulationTrap:, that invokes 
the relevant closure. Once the closure has completed, the processor simulator is advanced to the next pc and simulation resumes. In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM. Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation portable and processor independent.

>>>>

\paragraph{Production memory layout.}Figure \ref{fig:SimulatedMemory} describes, from a high-level perspective, the memory used by \openSmalltalkVM at runtime and the simulated counter-part. Let us detail briefly the memory used by the production VM, on the top of the figure. On the left side, which is usually the low addresses, we can first find Text, the memory section holding the native code of the VM (the compiled code of the interpreter, the memory manager and the JIT itself, but not the code compiled by the JIT). Then, Data holds initialized and uninitialized data, mainly the C variables used globally in the VM.

@@

All the Slang code is implemented in multiple Smalltalk classes, to organise the code and to add modularity through polymorphism. For example, the \texttt{AbstractCompactor} class has two subclasses, one implementing a sweep algorithm and the other a compact one. For production, at Slang-to-C compilation time, all the code is compiled in a single C file. All the modularity is removed, using the same example, the VM developer chooses at this moment if he wants to compile a VM with a sweep or compact algorithm. No polymorphism is available at runtime. However, since polymorphism is available in the simulator, it can be reused for debugging purposes. Still in our example, the \texttt{AbstractCompactor} class has also simulation specific subclasses. Such versions typically express additional constraints in the form of assertions which can be written in plain Smalltalk without restrictions to easily express complex constraints. They also keep specific values live so they can be accessed at debugging time.

@@

In addition to the interpreter simulator, simulating the JIT requires to simulate the execution of the native code it generates. The JIT itself is written in Slang and simulated with the Smalltalk execution model. To simulate the machine code, the start of the byte array representing the memory holds, when the JIT is enabled, the machine code generated at runtime.

Bindings to processor simulator libraries (Bochs for x86 and x64, Skyeye for ARMv6, Smalltalk code for 32-bit MIPSLE) were implemented so that the machine code can be executed safely. All but the MIPSLE simulators are accessed through %Smalltalk primitives.  These are effectively a combination of an 
FFI calls and surrounding glue code.
% to invoke foreign code plus Smalltalk code that is executed if (the code invoked by) the primitive returns an error. 
The next paragraphs describe in detail how this scheme is used to ensure that simulation and production machine code are as similar as possible and how simulated machine code is interfaced to the rest of the simulation. %, and the debugging facilities for machine code that are available.

@@

A key requirement to enable effective development of the JIT is the code generated during simulation to be as close as possible to the code generated by the production JIT.  The code is of course not bit-identical; the address space is laid out very differently in memory to the simulation's byte array and the simulator is not subject to Address Space Layout Randomisation. But we benefit from the code generated under simulation being otherwise identical to that generated in the production VM and we go to some effort to achieve this.  The technique we use, involving illegal addresses, also allows us to interface with machine code the Smalltalk objects that comprise the Simulation, the interpreter, memory manager and JIT, and variables within them. To explain this let's consider how the interpreter references a stack frame.

The interpreter has framePointer, stackPointer and instructionPointer variables used to interpret code. When machine code wants to enter the runtime, for example to have the runtime look up a send on a cache miss, or to invoke the scavenger when eden is full, it needs to record the current stack frame, writing the native stack and frame pointers into the stackPointer and framePointer variables, and then to call the relevant routine. The runtime can then access the current machine code stack frame though these variables. In the production VM, the stackPointer and framePointer variables are C variables at known addresses in Data, and routines in the runtime have been compiled from C and exit in Text. Referencing these variables and invoking runtime routines from machine code is therefore straight-forward.  But in the simulation they are inaccessible. First of all they are instance variables inside Smalltalk objects, objects that have strong encapsulation, accessible only through messages, and second of all they are alongside the simulation's byte array, not within it, therefore the processor simulator has no access to these variables.

%NOTE: I almost only commented part of what Eliot wrote but did not change anything. I basically removed implementation specific details.
The JIT maintains a set of Dictionaries (hash maps) that map integers representing addresses to closures that get or set variables in the simulation objects, or invoke runtime routines, in both cases by sending messages to the simulation objects. As code is JITted and specific variables and runtime routines are referenced, the JIT manufactures a unique illegal address to reference the variable or routine, and uses this as the key in the relevant map, creating and storing the relevant closure accessor as the value.  When the processor simulator attempts to execute an instruction containing one of these illegal addresses, which will be some variety of read, write, call, jump or return instruction\footnote{activations of JITted machine code methods return to calling interpreter frames by returning to a trampoline that transfers control to the interpreter; hence the return address of a machine code activation above an interpreter frame has this trampoline as its return address}, it raises a ProcessorException.  %Our processor simulators are invoked through a "primitive" interface, much like an FFI call, and the attempt to raise an exeption is caught by wrapper code that returns an error result from the primitve invocation; in Smalltalk terminology the primitive fails.  The primitive failure code, Smalltalk code in the method that contains the primitive invocation, then analyses the offending instruction and constructs a Smalltalk ProcessorException object.  This Smalltalk exception has a type (\#read, \#write, \#jump, \#call), the address of the memory location being read, written, jumped to or called, a program counter (pc) holding the instruction's address, a next pc, and an accessor for the register, if any, that is the target of the read or the source of the write.  
The processor simulator is invoked in the context of an exception handler for such exceptions which invokes %a Smalltalk method, handleSimulationTrap:, that invokes 
the relevant closure. Once the closure has completed, the processor simulator is advanced to the next pc and simulation resumes. In this way, machine code generated in simulation can access arbitrary simulation objects while remaining essentially identical to the generated machine code in the production VM. Further, ProcessorException, and a handful of other methods implemented by a processor simulator makes the interfacing of machine code to the Smalltalk part of the simulation portable and processor independent.
